---
title: "R Notebook"
output: html_notebook
---

Initially, we'd taken the peptide level output from PD. However, in the previous notebook, we demonstrated that these quantifications lead to ratio compression due to interference/co-isolation. Therefore, we'd like to start from the PSM-level quantification and manually filter on PSM SN and Interference


```{r}
source('../../CamProt_R/Utility.R')
library(tidyverse)
library(Hmisc)
```


```{r}
samples_inf <- '../raw/samples.tsv'
```
 
Read in the PSM level data
```{r}
total_psm_raw <- "../raw/Total_Arsenide_TMT_PSMs.txt"


total_psm <- parse_features(total_psm_raw, TMT=TRUE, filter_crap=TRUE, filter_associated_crap=TRUE, level='PSM')
total_psm_res <- total_psm %>% filter(is.finite(Isolation.Interference.in.Percent)) %>%
    makeMSNSet(samples_inf, level="PSM", ab_col_ix=2) 
```

```{r}
rbp_psm_raw <- "../raw/RBP_Arsenide_TMT_PSMs.txt"

rbp_psm <- parse_features(rbp_psm_raw, TMT=TRUE, filter_crap=TRUE, filter_associated_crap=TRUE, level='PSM')
rbp_psm_res <- rbp_psm %>% filter(is.finite(Isolation.Interference.in.Percent)) %>%
    makeMSNSet(samples_inf, level="PSM", ab_col_ix=2) 
```

Function to fitler PSM and aggregate to protein
```{r}
PSMToProtein <- function(
  raw_res, # the PSM level dataframe

  SN_threshold=5, # Signal:noise threshold to retain PSM quantification values
  intensity_filter=0, # Minimum intensity value, e.g 2^2.25
  interference_threshold=50, # Maximum interference
  mod_col="Modifications" # column to identify the modifications for grouping PSM into pep seq + mod
){

  # Remove PSMs with Interference above threshold
  raw_res_int <- raw_res[fData(raw_res)$Isolation.Interference.in.Percent<=interference_threshold,]
  
  # Remove PSMs with Signal:Noise (SN) below threshold
  raw_res_sn <- raw_res_int[fData(raw_res_int)$Average.Reporter.SN>=SN_threshold,]
  
  # Remove low intensity estimates
  exprs(raw_res_sn)[exprs(raw_res_sn)<intensity_filter] <- NA
  
  # Aggregate to unique peptide sequences
  agg_pep <- suppressWarnings(suppressMessages(agg_to_peptides(raw_res_sn)))
  invisible(agg_pep)}



  
```

```{r}
int_thresholds <- c(0, 10, 50)
pep_quants <- vector("list", length=length(int_thresholds))
names(pep_quants) <- sprintf('Int_%s', int_thresholds)

for(int in int_thresholds){
  threshold_quants <-vector("list", 2)
  names(threshold_quants) <- c('total', 'rbp')
  
  threshold_quants[['total']] <- PSMToProtein(
    total_psm_res, SN_threshold=10, interference_threshold=int)
  threshold_quants[['rbp']] <- PSMToProtein(
    rbp_psm_res, SN_threshold=10, interference_threshold=int)
    
  pep_quants[[sprintf('Int_%s', int)]] <- threshold_quants
}
```

```{r}
pep_quants$Int_0$total %>% plotMissing()
```

```{r}
plotMissing(total_psm_res)
plotMissing(rbp_psm_res)
```

```{r}
pep_quants %>% lapply(function(x){
  plotMissing(x$total)
  plotMissing(x$rbp)
})
```



```{r}
prot_quants <- pep_quants %>% lapply(function(x){
  
  # Aggregate to unique protein

  threshold_quants <-vector("list", 2)
  names(threshold_quants) <- c('total', 'rbp')
  
  threshold_quants$total <- suppressWarnings(suppressMessages(
    agg_to_protein(x$total, protein_col="Master.Protein.Accessions",))) %>%
    centerNormalise()

  threshold_quants$rbp <- suppressWarnings(suppressMessages(
    agg_to_protein(x$rbp, protein_col="Master.Protein.Accessions",))) %>%
    centerNormalise()
  
  return(threshold_quants)
  
})

  

```

```{r}

prot_quants %>% lapply(function(x){
  plotMissing(x$total)
  plotMissing(x$rbp)
})
```
```{r}
saveRDS(prot_quants, '../results/prot_quants_thresholds.rds')
```

