---
title: "05: Power simulations"
author: "Mariavittoria Pizzinga, Eneko Villanueva,Rayner Queiroz, Manasa Ramakrishna, Tom Smith"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  pdf_document:
    toc: yes
    toc_depth: '3'
  html_document:
    toc: yes
    toc_depth: 3
    toc_float: yes
editor_options: 
  chunk_output_type: inline
---
```{r global_options, echo=FALSE}
knitr::opts_chunk$set(fig.width=8,fig.height=6,warning=FALSE, message=FALSE,tidy = TRUE,tidy.opts=list(width.cutoff=50))
```
## 1. Introduction 
Here we simulate some datasets to estimate the number of replicates required to reach a reasonable power to detect change in RNA binding

```{r 00_Libraries-functions-needed, eval=T, echo=F, warning=F}

source("../../CamProt_R/Utility.R")
library(tidyverse)
library(reshape2)
library(dplyr)

# -----------------------------  
# Function	: plotCVs
# -----------------------------  
plotCVs <- function(obj){
  tmp <- exprs(obj)
  colnames(tmp) <- pData(obj)$Sample_name
  CVs <- melt(tmp) %>% separate(Var2, into=c("Condition", "Replicate")) %>%
    group_by(Var1, Condition) %>%
    dplyr::mutate("CV"=sd(2^value)/mean(2^value), "CV_log"=sd(value)/mean(value))
  
  p <- ggplot(CVs, aes(x=CV, colour=Condition)) + geom_density() + my_theme
  return(list("p"=p, "CVs"=CVs))
}

# -----------------------------  
# Function	: simulateValues
# -----------------------------  
simulateValues <- function(obj, median_cv, ix, max_reps, n_sims){
  simulated_values <- sapply(obj, FUN = function(x) rnorm(max_reps, x, x*median_cv))
  colnames(simulated_values) <- seq(ix, ix+(n_sims-1))
  rownames(simulated_values) <- 1:max_reps
  simulated_values <- melt(simulated_values)
  colnames(simulated_values) <- c("Replicate", "Protein_Ix", "Intensity")
  
  return(simulated_values)
}

# -----------------------------  
# Function	: getSimulatedDF
# -----------------------------  
getSimulatedDF <- function(change_range=seq(-2,2,0.2), median_total_CV=0.02, median_oops_CV=0.02){
  rows <- NULL
  ix = 1
  for (log2_diff_size in 2^change_range){
    total_abundance_c <- sample(total_abundances, n_sims, replace=TRUE)
    simulated_values_total_c <- simulateValues(total_abundance_c, median_total_CV, ix, max_reps, n_sims)
    simulated_values_total_c$Type <- "Total"
    simulated_values_total_c$Condition <- "Control"
  
    changes <- rnorm(n_sims, mean=1, sd=0.1)
    total_abundance_t <- total_abundance_c * changes
    simulated_values_total_t <- simulateValues(total_abundance_t, median_total_CV, ix, max_reps, n_sims)
    simulated_values_total_t$Type <- "Total"
    simulated_values_total_t$Condition <- "Treatment"
    
    oops_abundance_c <- sample(oops_abundances, n_sims, replace=TRUE)
    simulated_values_oops_c <- simulateValues(oops_abundance_c, median_oops_CV, ix, max_reps, n_sims) 
    simulated_values_oops_c$Type <- "OOPS"
    simulated_values_oops_c$Condition <- "Control"
    
    oops_changes <- changes * log2_diff_size
    oops_abundance_t <- oops_abundance_c * oops_changes
    simulated_values_oops_t <- simulateValues(oops_abundance_t, median_oops_CV, ix, max_reps, n_sims)
    simulated_values_oops_t$Type <- "OOPS"
    simulated_values_oops_t$Condition <- "Treatment"
  
      
    simulated_values <- rbind(simulated_values_total_c, simulated_values_total_t,
                              simulated_values_oops_c, simulated_values_oops_t)
    
    simulated_values$diff <- log2_diff_size
    if(nrow(simulated_values[is.na(simulated_values$Intensity),])>0){
      print(oops_changes)
      print(total_abundance_c)
      print(total_abundance_t)
      print(oops_abundance_c)
      print(oops_abundance_t)
      stop()
    }
    
    rows[[ix]] <- simulated_values
    ix <- ix + n_sims
  }
  
  simulated_df <- do.call("rbind", rows)
  
  simulated_df$Condition <- factor(simulated_df$Condition, levels=c("Control", "Treatment"))
  simulated_df$Type <- factor(simulated_df$Type, levels=c("Total", "OOPS"))
  simulated_df$Intensity <- log2(simulated_df$Intensity)
  
  dim(simulated_df)
  print(head(simulated_df))
  print(table(simulated_df$Condition, simulated_df$Type))
  

  return(simulated_df)
}

# -------------------------------------  
# Function	: testModels
# Aim       : To test linear models
# -------------------------------------

testModels <- function(obj, model="Intensity~Condition*Type", coeff_of_interest="ConditionTreatment:TypeOOPS"){

  fit <- obj %>% lm(formula=formula(model))
  #print(summary(fit))
  #print(obj)
  fit_values <- c(coef(summary(fit))[coeff_of_interest,],
                  summary(fit)$adj.r.squared)
  
  names(fit_values)[4:5] <- c("p_value", "adj_R_squared")
  
  return(fit_values)
}

# -------------------------------------  
# Function	: runLM
# Aim       : Run linear models
# -------------------------------------
runLM <- function(obj, model="Intensity~Condition*Type", coeff_of_interest="ConditionTreatment:TypeOOPS"){
   results <- obj %>%
    plyr::ddply("Protein_Ix", function(x) testModels(x, model=model, coeff_of_interest=coeff_of_interest))
  
  for(column in c("Estimate", "Std. Error", "t value", "adj_R_squared", "p_value")){
    results[[column]] <- as.numeric(results[[column]])
    }
   
  results$BH <- p.adjust(results$p_value, method="BH")
  
  return(results)
   
}

# -------------------------------------------------------------------------   
# Function	: runLM_multiple_reps
# Aim       : To run linear model using different number of replicates
# ------------------------------------------------------------------------- 
runLM_multiple_reps <- function(simulated_df, reps=2:10,
                                model="Intensity~Condition*Type",
                                coeff_of_interest="ConditionTreatment:TypeOOPS"){
  actual_changes <- simulated_df %>% dplyr::select(Protein_Ix, diff) %>% dplyr::distinct()
  simulated_lm_results <- NULL
  for (n_reps in reps){
    simulated_lm_results_rep <- simulated_df %>% dplyr::filter(Replicate<=n_reps) %>%
      runLM(., model, coeff_of_interest)  
    simulated_lm_results_rep$n_reps <- n_reps
    simulated_lm_results <- rbind(simulated_lm_results, simulated_lm_results_rep)
  }
  
  simulated_lm_results <- simulated_lm_results %>% merge(actual_changes, by="Protein_Ix")
  simulated_lm_results %>% arrange(desc(abs(diff))) %>% head() %>% print()
  
  return(simulated_lm_results)
}
# -------------------------------------------------------------------------   
# Function	: plot_results
# Aim       : To run linear model using different number of replicates
# ------------------------------------------------------------------------- 
plot_results <- function(obj){
  p <- obj  %>%
    group_by(n_reps, diff) %>%
    dplyr::summarise(power=sum(p_value<0.01)) %>%
    ggplot(aes(log2(diff), n_reps, fill=power)) +
    geom_tile(colour="grey50") +
    scale_x_continuous("Change in RNA binding (log2)", breaks=seq(-5,5,1)) +
    scale_y_continuous("Number of replicates", breaks=2:10) +
    my_theme
  
  print(p)
  
  p2 <- p + aes(fill=power>50)
  print(p2)
  
  p3 <- p + aes(fill=power>10)
  print(p3)
  
  invisible(p)
}

```
## 2. Reading in normalised protein level data

```{r 02_Read-in-normalised-data}
total_protein_quant <- readRDS("../results/total_as_res_pro_agg_norm")
oops_protein_quant <- readRDS("../results/rbp_as_res_pro_agg_norm")
```

A reminder of the % CV for the Total proteome and OOPS data

```{r 2b_Revisiting-CV-plots}
p_total <- plotCVs(total_protein_quant)
print(p_total$p + ggtitle("Total"))
#print(p_total + aes(CV_log))

p_oops <- plotCVs(oops_protein_quant)
print(p_oops$p+ ggtitle("OOPS"))
#print(p_oops + aes(CV_log))

```

## 3. Simulating a dataset
Make two simulation data sets, one with 2% CV, the other with observed CV (8% & 9% for Total and OOPS, respectively). We'll just use the median CV for each data set.

```{r 03_Simulating-datasets}

# Median CV calculation
median_total_CV <- median(p_total$CVs$CV)
median_oops_CV <- median(p_oops$CVs$CV)

total_abundances <- 2^p_total$CVs$value
oops_abundances <- 2^p_oops$CVs$value

n_sims <- 100
max_reps <- 10

simulated_df_2_perc <- getSimulatedDF()
simulated_df_observed_cv <- getSimulatedDF(median_total_CV=median_total_CV, median_oops_CV=median_oops_CV)

print(dim(simulated_df_2_perc))
print(dim(simulated_df_observed_cv))
```
Example simulated protein abundances at 2% CV and 8/9% CV (observed)

```{r 3a_Examples-of-simulated-data}

print(simulated_df_2_perc %>% filter(Protein_Ix==1))

simulated_df_2_perc %>% filter(Protein_Ix==1) %>%
  ggplot(aes(Condition, Intensity)) +
  geom_point() +
  my_theme +
  facet_wrap(~Type)

simulated_df_observed_cv %>% filter(Protein_Ix==1) %>%
  ggplot(aes(Condition, Intensity)) +
  geom_point() +
  my_theme +
  facet_wrap(~Type)
```
Run linear model on simulated data with differing numbers of replicates

```{r 3b_Linear-model-multireps-plot}

# Linear model on low CV data
simulated_lm_results_2_perc <- runLM_multiple_reps(simulated_df_2_perc)

# Linear model on observed CV data
simulated_lm_results_observed_cv <- runLM_multiple_reps(simulated_df_observed_cv)

# Plot of simulated data using observed CV
ggplot(simulated_lm_results_observed_cv, aes(adj_R_squared, colour=log2(diff), group=log2(diff))) +
  geom_density() +
  scale_colour_gradient2(high="blue", low="red", mid="black") +
  my_theme
```
```{r 3c_Plots-results-from-simulated-data}

# 2% CV simulated data
p <- plot_results(simulated_lm_results_2_perc)
ggsave("../plots/power_simulations_2_perc.png", p)

# Observed CV simulated data with filtering for p <0.01
simulated_lm_results_observed_cv %>% group_by(n_reps, diff) %>%
    dplyr::summarise(power=sum(p_value<0.01))

# Plot of the above
p <- plot_results(simulated_lm_results_observed_cv)
ggsave("../plots/power_simulations_observed_cvs.png", p)

```
## 4. Conclusions
So, with the observed CVs, it's hard to detect small changes in RNA binding (<2-fold). This is because we're looking at an interaction term so we require a lot more replicates to reach the same statistical power. Let's try an alternative approach where we just identify the changes in OOPS abundance and then classify the proteins according to their observed changes in total. Note, this is a less statisticall rigorous approach and the results will require a more cautious interpretation. 

```{r 4a_simulation-based-on-OOPS-only}

simulated_lm_results_2_perc_simple <- simulated_df_2_perc %>% filter(Type=="OOPS") %>%
  runLM_multiple_reps(reps=2:10, model="Intensity~Condition", coeff_of_interest="ConditionTreatment")

simulated_lm_results_observed_cv_simple <- simulated_df_observed_cv %>% filter(Type=="OOPS") %>%
  runLM_multiple_reps(reps=2:10, model="Intensity~Condition", coeff_of_interest="ConditionTreatment")


```


```{r 4b_Plotting-bases-on-OOPS-simulation}

# Based on 2% CV
plot_results(simulated_lm_results_2_perc_simple)

# Based on observed CV of 8-9%
plot_results(simulated_lm_results_observed_cv_simple)

```

