---
title: "R Notebook"
output: html_notebook
---


Here we simulate some datasets to estimate the number of replicates required to reach a reasonable power to detect change in RNA binding
```{r}
source("../../CamProt_R/Utility.R")
library(tidyverse)
library(reshape2)
```


```{r}
total_protein_quant <- readRDS("../results/total_res_pro_agg_norm")
oops_protein_quant <- readRDS("../results/rbp_res_pro_agg_norm")
```

A reminder of the % CV for the Total proteome and OOPS data
```{r}
plotCVs <- function(obj){
  tmp <- exprs(obj)
  colnames(tmp) <- pData(obj)$Sample_name
  CVs <- melt(tmp) %>% separate(Var2, into=c("Condition", "Replicate")) %>%
    group_by(Var1, Condition) %>%
    dplyr::mutate("CV"=sd(2^value)/mean(2^value), "CV_log"=sd(value)/mean(value))
  
  p <- ggplot(CVs, aes(x=CV, colour=Condition)) + geom_density() + my_theme
  return(list("p"=p, "CVs"=CVs))
}

p_total <- plotCVs(total_protein_quant)
print(p_total$p + ggtitle("Total"))
#print(p_total + aes(CV_log))

p_oops <- plotCVs(oops_protein_quant)
print(p_oops$p+ ggtitle("OOPS"))
#print(p_oops + aes(CV_log))
```
We'll just use the median CV for each data set
```{r}
median_total_CV <- median(p_total$CVs$CV)
median_oops_CV <- median(p_oops$CVs$CV)
```

Functions to simulate data
```{r}
total_abundances <- 2^p_total$CVs$value
oops_abundances <- 2^p_oops$CVs$value

n_sims <- 100
max_reps <- 10

simulateValues <- function(obj, median_cv, ix, max_reps, n_sims){
  simulated_values <- sapply(obj, FUN = function(x) rnorm(max_reps, x, x*median_cv))
  colnames(simulated_values) <- seq(ix, ix+(n_sims-1))
  rownames(simulated_values) <- 1:max_reps
  simulated_values <- melt(simulated_values)
  colnames(simulated_values) <- c("Replicate", "Protein_Ix", "Intensity")
  
  return(simulated_values)
}

getSimulatedDF <- function(change_range=seq(-2,2,0.2), median_total_CV=0.02, median_oops_CV=0.02){
  rows <- NULL
  ix = 1
  for (log2_diff_size in 2^change_range){
    total_abundance_c <- sample(total_abundances, n_sims, replace=TRUE)
    simulated_values_total_c <- simulateValues(total_abundance_c, median_total_CV, ix, max_reps, n_sims)
    simulated_values_total_c$Type <- "Total"
    simulated_values_total_c$Condition <- "Control"
  
    changes <- rnorm(n_sims, mean=1, sd=0.1)
    total_abundance_t <- total_abundance_c * changes
    simulated_values_total_t <- simulateValues(total_abundance_t, median_total_CV, ix, max_reps, n_sims)
    simulated_values_total_t$Type <- "Total"
    simulated_values_total_t$Condition <- "Treatment"
    
    oops_abundance_c <- sample(oops_abundances, n_sims, replace=TRUE)
    simulated_values_oops_c <- simulateValues(oops_abundance_c, median_oops_CV, ix, max_reps, n_sims) 
    simulated_values_oops_c$Type <- "OOPS"
    simulated_values_oops_c$Condition <- "Control"
    
    oops_changes <- changes * log2_diff_size
    oops_abundance_t <- oops_abundance_c * oops_changes
    simulated_values_oops_t <- simulateValues(oops_abundance_t, median_oops_CV, ix, max_reps, n_sims)
    simulated_values_oops_t$Type <- "OOPS"
    simulated_values_oops_t$Condition <- "Treatment"
  
      
    simulated_values <- rbind(simulated_values_total_c, simulated_values_total_t,
                              simulated_values_oops_c, simulated_values_oops_t)
    
    simulated_values$diff <- log2_diff_size
    if(nrow(simulated_values[is.na(simulated_values$Intensity),])>0){
      print(oops_changes)
      print(total_abundance_c)
      print(total_abundance_t)
      print(oops_abundance_c)
      print(oops_abundance_t)
      stop()
    }
    
    rows[[ix]] <- simulated_values
    ix <- ix + n_sims
  }
  
  simulated_df <- do.call("rbind", rows)
  
  simulated_df$Condition <- factor(simulated_df$Condition, levels=c("Control", "Treatment"))
  simulated_df$Type <- factor(simulated_df$Type, levels=c("Total", "OOPS"))
  simulated_df$Intensity <- log2(simulated_df$Intensity)
  
  dim(simulated_df)
  print(head(simulated_df))
  print(table(simulated_df$Condition, simulated_df$Type))
  

  return(simulated_df)
  }
```

Make two simulation data sets, one with 2% CV, the other with observed CV (8% & 9% for Total and OOPS, respectively)
```{r}
simulated_df_2_perc <- getSimulatedDF()
simulated_df_observed_cv <- getSimulatedDF(median_total_CV=median_total_CV, median_oops_CV=median_oops_CV)

print(dim(simulated_df_2_perc))
print(dim(simulated_df_observed_cv))
```

Functions to run linear models
```{r}
testModels <- function(obj, model="Intensity~Condition*Type", coeff_of_interest="ConditionTreatment:TypeOOPS"){

  fit <- obj %>% lm(formula=formula(model))
  #print(summary(fit))
  #print(obj)
  fit_values <- c(coef(summary(fit))[coeff_of_interest,],
                  summary(fit)$adj.r.squared)
  
  names(fit_values)[4:5] <- c("p_value", "adj_R_squared")
  
  return(fit_values)
}

runLM <- function(obj, model="Intensity~Condition*Type", coeff_of_interest="ConditionTreatment:TypeOOPS"){
   results <- obj %>%
    plyr::ddply("Protein_Ix", function(x) testModels(x, model=model, coeff_of_interest=coeff_of_interest))
  
  for(column in c("Estimate", "Std. Error", "t value", "adj_R_squared", "p_value")){
    results[[column]] <- as.numeric(results[[column]])
    }
   
  results$BH <- p.adjust(results$p_value, method="BH")
  
  return(results)
   
}

```

Example simulated protein abundances at 2% CV and 8/9% CV
```{r}
print(simulated_df_2_perc %>% filter(Protein_Ix==1))

simulated_df_2_perc %>% filter(Protein_Ix==1) %>%
  ggplot(aes(Condition, Intensity)) +
  geom_point() +
  my_theme +
  facet_wrap(~Type)

simulated_df_observed_cv %>% filter(Protein_Ix==1) %>%
  ggplot(aes(Condition, Intensity)) +
  geom_point() +
  my_theme +
  facet_wrap(~Type)
```
Function to run the linear models using 3-10 replicate
```{r}

runLM_multiple_reps <- function(simulated_df, reps=2:10,
                                model="Intensity~Condition*Type",
                                coeff_of_interest="ConditionTreatment:TypeOOPS"){
  actual_changes <- simulated_df %>% select(Protein_Ix, diff) %>% distinct()
  simulated_lm_results <- NULL
  for (n_reps in reps){
    simulated_lm_results_rep <- simulated_df %>% filter(Replicate<=n_reps) %>%
      runLM(., model, coeff_of_interest)  
    simulated_lm_results_rep$n_reps <- n_reps
    simulated_lm_results <- rbind(simulated_lm_results, simulated_lm_results_rep)
  }
  
  simulated_lm_results <- simulated_lm_results %>% merge(actual_changes, by="Protein_Ix")
  simulated_lm_results %>% arrange(desc(abs(diff))) %>% head() %>% print()
  
  return(simulated_lm_results)
}

simulated_lm_results_2_perc <- runLM_multiple_reps(simulated_df_2_perc)
simulated_lm_results_observed_cv <- runLM_multiple_reps(simulated_df_observed_cv)
```
```{r}
ggplot(simulated_lm_results_observed_cv, aes(adj_R_squared, colour=log2(diff), group=log2(diff))) +
  geom_density() +
  scale_colour_gradient2(high="blue", low="red", mid="black") +
  my_theme
```


Plot the results
```{r}

plot_results <- function(obj){
  p <- obj  %>%
    group_by(n_reps, diff) %>%
    dplyr::summarise(power=sum(p_value<0.01)) %>%
    ggplot(aes(log2(diff), n_reps, fill=power)) +
    geom_tile(colour="grey50") +
    scale_x_continuous("Change in RNA binding (log2)", breaks=seq(-5,5,1)) +
    scale_y_continuous("Number of replicates", breaks=2:10) +
    my_theme
  
  print(p)
  
  p2 <- p + aes(fill=power>50)
  print(p2)
  
  p3 <- p + aes(fill=power>10)
  print(p3)
  
  invisible(p)
}

p <- plot_results(simulated_lm_results_2_perc)
ggsave("../results/power_simulations_2_perc.png", p)
```

```{r}
simulated_lm_results_observed_cv %>% group_by(n_reps, diff) %>%
    dplyr::summarise(power=sum(p_value<0.01))

p <- plot_results(simulated_lm_results_observed_cv)
ggsave("../results/power_simulations_observed_cvs.png", p)

```

So, with the observed CVs, it's hard to detect small changes in RNA binding (<2-fold). This is because we're looking at an interaction term so we require a lot more replicates to reach the same statistical power. Let's try an alternative approach where we just identify the changes in OOPS abundance and then classify the proteins according to their observed changes in total. Note, this is a less statisticall rigorous approach and the results will require a more cautious interpretation. 

```{r}

simulated_lm_results_2_perc_simple <- simulated_df_2_perc %>% filter(Type=="OOPS") %>%
  runLM_multiple_reps(reps=2:10, model="Intensity~Condition", coeff_of_interest="ConditionTreatment")
simulated_lm_results_observed_cv_simple <- simulated_df_observed_cv %>% filter(Type=="OOPS") %>%
  runLM_multiple_reps(reps=2:10, model="Intensity~Condition", coeff_of_interest="ConditionTreatment")
#simulated_lm_results_observed_cv_simple <- runLM_multiple_reps(simulated_df_observed_cv, model="Intensity~Condition",
#                                                          coeff_of_interest="ConditionTreatment")
```
```{r}
simulated_df_2_perc %>% filter(Type=="OOPS")
```

```{r}
p <- plot_results(simulated_lm_results_2_perc_simple)
```

```{r}
p <- plot_results(simulated_lm_results_observed_cv_simple)
```

