---
title: "R Notebook"
output: html_notebook
---

Here, we take the results from limma and look for GO functional groups which show trends for concordant



```{r}
library(gage)

Ctrl.100uM <- readRDS("../results/Ctrl.100uM.rds")
Ctrl.400uM <- readRDS("../results/Ctrl.400uM.rds")
Ctrl.400uM[,45:52] %>% tibble::rownames_to_column() %>% arrange(desc(logFC))
Ctrl.400uM[,45:52] %>% arrange(desc(logFC)) %>% tail()

Ctrl.400uM
```

```{r}
human_go <- readRDS("../shared_files/h_sapiens_go_full.rds")

translation_init_activity <- human_go %>% filter(GO.ID=='GO:0003743') %>% pull(UNIPROTKB)
translation_elong_activity <- human_go %>% filter(GO.ID=='GO:0003746') %>% pull(UNIPROTKB)
translation_term_activity <- human_go %>% filter(GO.ID=='GO:0008079') %>% pull(UNIPROTKB)
tRNA_AA <- human_go %>% filter(GO.ID=='GO:0004812') %>% pull(UNIPROTKB)
translocon <- human_go %>% filter(GO.ID=='GO:0006616') %>% pull(UNIPROTKB)

gotoi <- list(translation_init_activity, translation_elong_activity,
              translation_term_activity, tRNA_AA, translocon )
names(gotoi) <- c('GO:0003743', 'GO:0003746', 'GO:0008079', 'GO:0004812', 'GO:0006616')
print(gotoi)

all_go_terms <- human_go %>% filter(UNIPROTKB %in% rownames(Ctrl.400uM)) %>%
  pull(TERM) %>% unique()

all_go <- vector("list", length=length(all_go_terms))
names(all_go) <- all_go_terms

for(x in all_go_terms){
  all_go[[x]] <- human_go %>% filter(TERM==x) %>% pull(UNIPROTKB)
}

print(head(all_go,1))
```


```{r}

ranks <- rev(sort(Ctrl.100uM$logFC))
names(ranks) <- rownames(Ctrl.100uM)
head(ranks)

barplot(ranks)

fgseaRes <- fgsea(all_go, ranks, minSize=15, maxSize = 500, nperm=1000)
head(fgseaRes[order(pval, -abs(NES)), ], n=10)

gotoi %>% lapply(function(x){
  plotEnrichment(x, ranks)
})

head(fgseaRes[order(pval, -abs(NES)), ], n=10)$pathway %>% lapply(function(x){
  plotEnrichment(all_go[[x]], ranks) + ggtitle(x)
})

plot(ranks)
```
```{r}
fgseaRes %>% arrange(pval) %>% filter(grepl("Ribosome", pathway, ignore.case=TRUE))
fgseaRes %>% arrange(pval) %>% filter(grepl("Endoplasmic", pathway, ignore.case=TRUE))
fgseaRes %>% arrange(pval) %>% filter(grepl("Translation", pathway, ignore.case=TRUE))
```

```{r}



plot_foi_trends <- function(obj, foi){
  foi_proteins <- all_go[[foi]]
    
  p <- obj %>%
    mutate(sig=ifelse(adj.P.Val<0.1, "sig.", "not sig.")) %>% # add "sig" column
    mutate(foi=Master.Protein.Accessions %in% foi_proteins) %>%
    arrange(foi) %>%
    ggplot(aes(logFC, -log10(P.Value), colour=foi, shape=sig)) +
    geom_point() +
    scale_colour_manual(values=c("black", cbPalette[6]), name="FOI") +# manually adjust colours
    scale_size_manual(values=c(1,2), name="FOI")  +# manually adjust sizes
    scale_size_discrete(name="Sig. (10% FDR)") +
    ggtitle(foi)
  
  print(p)
  
  print(plotEnrichment(all_go[[foi]], ranks) +
  ggtitle(foi))
  
  invisible(p)
}


plot_foi_trends(Ctrl.100uM, 'regulation of translation')
plot_foi_trends(Ctrl.100uM, 'tRNA aminoacylation for protein translation')
plot_foi_trends(Ctrl.100uM, 'structural constituent of ribosome')
plot_foi_trends(Ctrl.100uM, 'endoplasmic reticulum unfolded protein response')


```


```{r}
topUp <- fgseaRes %>% 
    filter(ES > 0) %>% 
    top_n(10, wt=-pval)
topDown <- fgseaRes %>% 
    filter(ES < 0) %>% 
    top_n(10, wt=-pval)
topPathways <- bind_rows(topUp, topDown) %>% 
    arrange(-ES)
x <- plotGseaTable(all_go[topPathways$pathway], 
                   gseaParam = 0.5,
              ranks, 
              fgseaRes)
str(x)
```


Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file). 

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

