---
title: "R Notebook"
output: html_notebook
---

We have observed some evidence of ratio compression for PSMs with high interference and/or low SN. Now, let's see how these thresholds impact the fold-differences between samples/conditions. For this, we will use the parse

```{r}
source('../../CamProt_R/Utility.R')
library(tidyverse)
library(Hmisc)
```

```{r}
samples_inf <- '../raw/samples.tsv'
```
 
Read in the PSM level data
```{r}
total_psm_raw <- "../raw/Total_Arsenide_TMT_PSMs.txt"
total_psm <- parse_features(total_psm_raw, TMT=TRUE, filter_crap=TRUE, filter_associated_crap=TRUE, level='PSM')
total_psm_res <- total_psm %>% filter(is.finite(Isolation.Interference.in.Percent)) %>%
    makeMSNSet(samples_inf, level="PSM", ab_col_ix=2) 
```

Function to fitler PSM and aggregate to protein
```{r}
PSMToProtein <- function(
  raw_res, # the PSM level dataframe
  master_protein_col="Master.Protein.Accessions", # master protein column name
  SN_threshold=5, # Signal:noise threshold to retain PSM quantification values
  intensity_filter=0, # Minimum intensity value, e.g 2^2.25
  interference_threshold=50, # Maximum interference
  mod_col="Modifications" # column to identify the modifications for grouping PSM into pep seq + mod
){

  # Remove PSMs with Interference above threshold
  raw_res_int <- raw_res[fData(raw_res)$Isolation.Interference.in.Percent<=interference_threshold,]
  
  # Remove PSMs with Signal:Noise (SN) below threshold
  raw_res_sn <- raw_res_int[fData(raw_res_int)$Average.Reporter.SN>=SN_threshold,]
  
  # Remove low intensity estimates
  exprs(raw_res_sn)[exprs(raw_res_sn)<intensity_filter] <- NA
  
  # Aggregate to unique peptide sequences
  agg_pep <- suppressWarnings(suppressMessages(agg_to_peptides(raw_res_sn)))
  
  # Aggregate to unique protein
  agg_prot <- suppressWarnings(suppressMessages(agg_to_protein(agg_pep, master_protein_col))) %>%
    centerNormalise()
  
  invisible(agg_prot)
}
  
```

```{r}
x <- PSMToProtein(total_psm_res, SN_threshold=5)
```

```{r}
dim(x)
```
Function to obtain the fold changes for each comparison and plot to check this is working
```{r}
getFoldChanges <- function(obj, group1, group2){

  fc <- obj %>% exprs() %>%
  apply(MARGIN=1, FUN=function(x){
    g1=mean(x[group1], na.rm=TRUE)
    g2=mean(x[group2], na.rm=TRUE)
    
    if(any(is.na(c(g1, g2)))){
      return(NA)
      } 
    else{
      return(g2/g1)
      }
  
  })
  
  invisible(log2(fc))
}


getAllFoldChanges <- function(obj){
  
  g1g2 <- getFoldChanges(obj, 1:3, 4:7) %>% data.frame() %>%
    tibble::rownames_to_column('protein') %>%
    mutate(comp='Ctrl:Arsenite100')
  
  g1g3 <- getFoldChanges(obj, 1:3, 8:10) %>% data.frame() %>%
    tibble::rownames_to_column('protein') %>%
    mutate(comp='Ctrl:Arsenite400')
  
  g2g3 <- getFoldChanges(obj, 4:7, 8:10) %>% data.frame() %>%
    tibble::rownames_to_column('protein') %>%
    mutate(comp='Arsenite100:Arsenite400')
  
  results <- do.call('rbind', list(g1g2, g1g3, g2g3))
  colnames(results)[2] <- 'fc'
  
  return(results)
}

getAllFoldChanges(x) %>% ggplot(aes(fc, colour=comp)) + geom_density() + my_theme
```

Get fold changes for each possible combination
```{r}
fc_dfs <- NULL
for(sn in c(0,5,10,100)){
  for(int in c(0,5,10,50,100)){
    x <- PSMToProtein(total_psm_res, SN_threshold=sn, interference_threshold=int, intensity_filter=0)
    fc <- getAllFoldChanges(x)
    fc$sn <- sn
    fc$int <- int
    fc_dfs[[sprintf('%s_%s', sn, int)]] <- fc
  }
    
}
```

How many proteins per combination?
```{r}
n_proteins <- fc_dfs %>% lapply(function(x){
  length(unique(x$protein))
})

n_proteins <- n_proteins %>% data.frame() %>% t() %>% data.frame() %>% 
  tibble::rownames_to_column() %>%
  mutate(parameters=gsub("X", "", rowname)) %>%
  separate(parameters, into=c("SN", "interference")) %>% select(-rowname)
colnames(n_proteins)[1] <- 'proteins'

n_proteins %>% filter(SN==100) %>% select(-SN)
n_proteins %>% filter(interference==0) %>% select(-interference)
```

Plot all fold changes. Testing aethestics for plotting
```{r}

do.call('rbind', fc_dfs) %>%
  ggplot(aes(fc, colour=factor(int))) + geom_density() + my_theme +
  facet_grid(comp~sn, scales='free_y')

do.call('rbind', fc_dfs) %>%
  ggplot(aes(fc, colour=factor(sn))) + geom_density() + my_theme +
  facet_grid(comp~int, scales='free_y')
```

We're only really interested in subsets of proteins. Let's define these via GO annotations
```{r}
human_go <- readRDS('../shared_files/h_sapiens_go_full.rds')

translation_init_activity <- human_go %>% filter(GO.ID=='GO:0003743') %>% pull(UNIPROTKB)
tRNA_AA <- human_go %>% filter(GO.ID=='GO:0004812') %>% pull(UNIPROTKB)


ribosome_structural_component <- human_go %>% filter(GO.ID=='GO:0003735') %>% pull(UNIPROTKB)
cytosol <- human_go %>% filter(GO.ID=='GO:0005829') %>% pull(UNIPROTKB)
ribosome_structural_component_cytosol <- intersect(ribosome_structural_component, cytosol)
print(length(ribosome_structural_component))
print(length(cytosol))
print(length(ribosome_structural_component_cytosol))
```

```{r, fig.width=10, fig.height=10}
do.call('rbind', fc_dfs) %>%
  filter(protein %in% ribosome_structural_component_cytosol) %>%
  ggplot(aes(fc, colour=factor(sn))) + geom_density() + my_theme +
  facet_grid(comp~int, scales='free_y')

do.call('rbind', fc_dfs) %>%
  filter(protein %in% ribosome_structural_component_cytosol) %>%
  ggplot(aes(fc, colour=factor(int))) + geom_density() + my_theme +
  facet_grid(comp~sn, scales='free_y')
```
```{r, fig.width=6, fig.height=6}
sn_100_int_0_proteins <- unique(fc_dfs[["100_0"]]$protein)
print(length(sn_100_int_0_proteins))

do.call('rbind', fc_dfs) %>%
  filter(protein %in% ribosome_structural_component_cytosol, 
         int==0, protein %in% sn_100_int_0_proteins) %>%
  ggplot(aes(fc, colour=factor(sn))) +
  geom_density() + my_theme +
  geom_vline(xintercept=0, linetype=2, colour='grey50', size=0.5) +
  facet_grid(comp~., scales='free_y') +
  theme(strip.text=element_text(size=10)) +
  scale_colour_manual(values=c("black", cbPalette[c(2:4,7)]),name='SN') +
  ylab("Density") +
  scale_x_continuous(name="Fold change (log2)")
```

```{r, fig.width=6, fig.height=6}
sn_100_int_0_proteins <- unique(fc_dfs[["100_0"]]$protein)
print(length(sn_100_int_0_proteins))

do.call('rbind', fc_dfs) %>%
  filter(protein %in% ribosome_structural_component_cytosol, 
         sn==100, protein %in% sn_100_int_0_proteins) %>%
  ggplot(aes(fc, colour=factor(int))) +
  geom_density() + my_theme +
  geom_vline(xintercept=0, linetype=2, colour='grey50', size=0.5) +
  facet_grid(comp~., scales='free_y') +
  theme(strip.text=element_text(size=10)) +
  scale_colour_manual(values=c("black", cbPalette[c(2:4,7)]),name='Interference') +
  ylab("Density") +
  scale_x_continuous(name="Fold change (log2)")
```
```{r}
do.call('rbind', fc_dfs) %>%
  filter(protein %in% translation_init_activity, sn==100,
         protein %in% sn_100_int_0_proteins) %>%
  ggplot(aes(fc, colour=factor(int))) +
  geom_density() + my_theme +
  geom_vline(xintercept=0, linetype=2, colour='grey50', size=0.5) +
  facet_grid(comp~., scales='free_y') +
  theme(strip.text=element_text(size=10)) +
  scale_colour_manual(values=c("black", cbPalette[c(2:4,7)]),name='Interference') +
  ylab("Density") +
  scale_x_continuous(name="Fold change (log2)")

```
```{r}
do.call('rbind', fc_dfs) %>%
  filter(protein %in% tRNA_AA, sn==100,
         protein %in% sn_100_int_0_proteins) %>%
  ggplot(aes(fc, colour=factor(int))) +
  geom_density() + my_theme +
  geom_vline(xintercept=0, linetype=2, colour='grey50', size=0.5) +
  facet_grid(comp~., scales='free_y') +
  theme(strip.text=element_text(size=10)) +
  scale_colour_manual(values=c("black", cbPalette[c(2:4,7)]),name='Interference') +
  ylab("Density") +
  scale_x_continuous(name="Fold change (log2)")


```

